Over Twenty Years of Virtual Machine Development and Debugging Through Simulation

### 2018-08-14 paper.tex
kept second pass

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid. To work around this problem, we implemented a tool called \emph{In-image compilation}. In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly\footnote{Note that this apply for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.}. HERE Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates. To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To work around this, we built a facade, which is started on the compiled method and only includes the subset of objects required by the JIT to generate the machine code.

@@

\paragraph{Performance.}
The first limitation is due to the simulation performance. The interpreter-only is simulator is around 200 times slower to execute code than the normal VM. With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM. This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator. Bugs in the jitted code are worst. Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs. In general, once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes. The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds. If the bug is unrelated to the JIT, the interpreter-only simulator can be used and it is a little bit quicker to execute code.

\paragraph{Calls to external code.}
Although most of the GC and JIT development and bug resolution can be done in the simulator, specific developments and bugs cannot. Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated. For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation. However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT cannot be simulated (Access to C variables, OS, etc.).

The main limitation we have is with Foreign Function Interfaces (FFI). Most bugs we have with FFI are due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects. Such bugs cannot be debugged, so far, with our simulation infrastructure and we have to rely on gdb/lldb.

@@

Recently, to evaluate new garbage collection algorithm against standard algorithm, we implemented a Mark-Sweep in addition to the existing Mark-Compact for old space garbage collection. The whole implementation was done in the simulator, and only when it was working there, we compiled it to C and it was working out of the box. To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.

>>>>

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid. To work around this problem, we implemented a tool called \emph{In-image compilation}. In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code and display the disassembly\footnote{Note that this apply for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.}.

The JIT basically generates a sequence of machine instructions for a given pattern of bytecodes. We will use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and  
a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates. To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To work around this, we built a facade, which includes mock addresses for all the object the JIT may require to generate the machine code of a given method.

@@

The simulator has several limitations. Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production. However, we have been using this infrastructure since 1995, and this kind of bugs are really rare and usually easy to fix. We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot be simulated.

\paragraph{Performance.}
The first limitation is due to the simulation performance. The interpreter-only is simulator is around 200 times slower to execute code than the normal VM. With the JIT and processor simulation enabled, without specific debugging options such as conditional breakpoints in between machine instructions, simulation drops to around 500 times slower than the normal VM. We usually enabled conditional breakpoints in-between machine instructions only when we reach a point in the simulator where the bug is about to happen since it is way slower, so the overall performance in this context is not really relevant.

This means for example that if a GC bug happens in an application 15 minutes after start-up, it will take 50 hours to reproduce in the interpreter-only simulator. Bugs in the jitted code are worse. Fortunately, we work around this problem by using snapshots and the interpreter-only simulator for GC bugs. In general, once we are able to reproduce a bug in the production VM, we try to snapshot the runtime just before it crashes. The VM simulator can then be started just before the crash and the debugging tools can be used after only several dozens of seconds. If the bug is unrelated to the JIT, the interpreter-only simulator can also be used and it is a little bit quicker to execute code.

\paragraph{Calls to external code.}
Although most of the GC and JIT development and debugging can be done in the simulator, specific tasks cannot be done this way. Basically, any calls outside of the machine code generated by the JIT and the Slang code cannot be simulated. For specific small parts of the VM, such as file management, we extended the simulator, effectively duplicating the code base with the C code, to support those features in simulation. However, there is no solution in the general case: we cannot afford to simulate both the compiled C code and the jitted code on the processor simulator, that would be horribly slow, and specific behaviors in the machine code not present in the code generated by the JIT can hardly be simulated (Access to C variables, OS variables, etc.).

The main limitation we have right now is with Foreign Function Interfaces (FFI). We have a significant amount of bugs in FFI, often due to specific interaction between call-backs, low-level assembly FFI specific glue code and moving objects. Such bugs cannot be debugged with our simulation infrastructure so far and we have to rely on gdb/lldb.

@@

HERE
Recently, to evaluate new garbage collection algorithm against standard algorithm, we implemented a Mark-Sweep in addition to the existing Mark-Compact for old space garbage collection. The whole implementation was done in the simulator, and only when it was working there, we compiled it to C and it was working out of the box. To describe the implementation process, we need to discuss briefly first the assertion levels in the VM. Then we will show how we debugged the algorithm.
